\pkg{dataMaid}: your assistant for data documentation in \proglang{R}

### 2017-10-18 latex/article_vol3.tex
New version of article, halfway done.

%% almost as usual
\author{Anne H. Petersen\\Biostatistics\\Department of Public
  Health\\University of Copenhagen \And Claus Thorn Ekstr\o m\\Biostatistics\\Department of Public
  Health\\University of Copenhagen}
\title{\pkg{dataMaid}: your assistant for data cleaning in \proglang{R}}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Anne H. Petersen, Claus Thorn Ekstr\o m} %% comma-separated
\Plaintitle{{dataMaid}: your assistant for data cleaning in R} %% without formatting
\Shorttitle{\pkg{dataMaid}: your data cleaning personal assistant in R} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{Data cleaning and -validation are the first steps in any
  data analysis, as the validity of the conclusions from the analysis
  hinges on the quality of the input data. Mistakes in the data can
  arise for any number of reasons, including erroneous codings,
  malfunctioning measurement equipment, and inconsistent data generation
  manuals.  Ideally, a human investigator should go
  through each variable in the dataset and look for potential errors
  --- both in input values and codings --- but that process can be very
  time-consuming, expensive and error-prone in itself.

  We describe an \proglang{R} package, \pkg{dataMaid}, which
  implements an extensive and customizeable suite of quality
  assessment tools that can be applied to a dataset in order to
  identify potential problems in its variables. The results can be
  presented in an auto-generated, non-technical, stand-alone overview
  document intended to be perused by an investigator with an
  understanding of the variables in the data, but not necessarily
  knowledge of \proglang{R}. Thereby, \pkg{dataMaid} aids the dialogue
  between data analysts and field experts, while also providing easy
  documentation of reproducible data cleaning steps and data quality
  control. Moreover, the \pkg{dataMaid} solution changes the data
  cleaning process from the usual ad hoc approach to a systematic,
  well-documented endeavor.  \pkg{dataMaid} also provides a suite of
  more typical \proglang{R} tools for interactive data quality
  assessment and -cleaning, where the data inspections all live and
  die within the \proglang{R} console.

  % The \pkg{dataMaid} package is designed to be easily extended with
  % custom user-created checks that are relevant in particular
  % situations. \hl{Already said that above. Either expand on it or
  % delete this.}
}
\Keywords{data cleaning, quality control, \proglang{R}}
\Plainkeywords{data cleaning, quality control, R} %% without formatting
%% at least one keyword must be supplied

@@

\section{Introduction}
Though data cleaning might be regarded as a somewhat tedious activity,
adequate data cleaning is crucial in any data analysis. With
ever-growing dataset sizes and complexities, statisticians and data
analysts find themselves spending a large portion of their time on
data cleaning and data wrangling. While a computer should never
make unsupervised decisions on what should be done to potential
errors in a dataset, it can still be an extremely useful tool in the
data cleaning process. Some errors can be tracked down and flagged by a
computer without further ado, while other types of errors need an analytic
context in order to be identified. Even in this latter case, well-designed
software can aid the process tremendously by providing the human
investigator with the
information needed for identifying issues.

A number of \proglang{R} packages made for such pre-analysis tools are
already available, including \pkg{janitor} \citep{janitor}, \pkg{plyr}
\citep{plyr}, \pkg{data.table} \citep{data.table}, \pkg{DataCombine}
\citep{DataCombine} , \pkg{validate} \citep{validate}, and
\pkg{DataExplorer} \citep{DataExplorer}. These packages focus on
different stages of the pre-analysis work, most of which are not
really data cleaning.  \pkg{janitor} provides tools for data import
with a particular emphasis on the challenges of getting neat data
frames from Microsoft Excel data files. \pkg{plyr}, \pkg{data.table}
and \pkg{DataCombine} go a few steps further by providing a wide array
of extremely powerful tools for data wrangling, including a number of
particularly useful functions for merging and working with very large
datasets. When it comes to actual data cleaning, however, the options
are fewer.  \pkg{validate} (and the similar packages \pkg{editrules}
\citep{editrules} and \pkg{deducorrect} \citep{deducorrect} from the
same authors) does offer a few tools for identifying and
auto-correcting errors in a dataset, but, as the name suggests, the
focus is on internal validity rather than general data cleaning. In
practice, this means that quite elegant tools for, e.g., linear
restraints among the variables in a dataset can be applied, but
looking for potentially miscoded missing values is not really
feasible. The main difference between these two challenges is the
direction in which the data is inspected: While linear constraints
work row-wise with no ambiguity, determining whether or not something
is a miscoded missing value often requires knowledge about the full
variable (e.g. range or data type), and thus it should be performed
column-wise. \pkg{validate} does not allow for user-defined extensions
of the latter type, thereby limiting its data cleaning potential.
Automatic data correction functions are also provided by
\pkg{validate} which we consider to be quite a dangerous cocktail: all
power is given to the the computer with no human supervision, and
investigators are less likely to make an active, case-specific choice
regarding the handling of the potential errors.  Finally, no tools
have been made to easily document exactly which checks and preliminary
results were used in the data cleaning process.

@@

The \pkg{dataMaid} approach to data cleaning and -quality assessment is
governed by two fundamental paradigms. First of all, there is no need
for data cleaning to be an ad hoc procedure. Often, we have a very
clear idea of what flags are raisable in a given dataset before we
look at it, as we were the ones to produce it in the first place. This
means that data cleaning can easily be a well-documented,
well-specified procedure. In order to aid this paradigm, \pkg{dataMaid}
provides easy-to-use, automated tools for data quality assessment in
\proglang{R} \citep{R} on which data cleaning decisions can be made. This quality
assessment is presented in an auto-generated overview document,
readable by data analysts and field experts alike, thereby also
contributing to an inter-field dialogue about the data at
hand. Oftentimes, e.g., distinguishing between faulty codings of a
numeric value and unusual, but correct, values requires
problem-specific expertise that might not be held by the data
analyst. Hopefully, having easy access to data descriptions through
\pkg{dataMaid} will help this necessary knowledge sharing.

@@

This manuscript is structured as follows: First, in Section
\ref{sec:usingdataMaid}, we introduce the main representative of the first
paradigm, namely the \code{clean()} function, which generates data
cleaning overview documents. In the \pkg{dataMaid} package, we have
provided a number of default cleaning steps that cover the data
cleaning challenges, we find to be most common. Next, in Section
\ref{sec:interactiveCleanR}, we present the interactive mode of \pkg{dataMaid}, as motivated
by the second paradigm above. But, as any data analyst knows,
every dataset is different, and some datasets might include problems
that cannot be detected by our data checking functions. In Section
\ref{sec:extending}, we turn to the question of how \pkg{dataMaid} extensions
can be made, such that they are integrable with the \code{clean()}
function and with the other tools available in \pkg{dataMaid}.  At last,
in Section \ref{sec:specificExamples}, we discuss a number of examples of
specific data cleaning challenges and how \pkg{dataMaid} can be used to
solve them.

@@

\section{Creating a data cleaning overview}
\label{sec:usingdataMaid}

The \code{clean()} function is the primary workhorse of \pkg{dataMaid} and
this is the only function needed if a standard battery
of tests are used to generate data cleaning summaries. The data
cleaning summary itself is an overview document, intended for reading
by humans, in either pdf or html format. Appendix \ref{sec:appendix1}
provides an example of a data cleaning output document, produced by
calling \code{clean()} on the dataset \code{toyData} available in
\pkg{dataMaid}. The first two pages of this data cleaning output are
shown in Figure~\ref{fig:example1}. \code{toyData} is a very
small ($15$ rows of $6$ variables), artificial dataset which was created to
illustrate the main capabilities of \pkg{dataMaid}. The following
commands load the dataset and produce the cleaning output:

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=2]{dataMaid_toyData.pdf}}
\frame{\includegraphics[width=7.5cm,page=3]{dataMaid_toyData.pdf}}
%\includepdf[pages={2}, pagecommand={}]{dataMaid_testData.pdf}
\end{center}
\caption{Example output from running \code{clean()} on the \code{toyData}
  dataset. First a summary of the full dataset is given and then
  type-dependent information on each variable is given in a table and
  a graph. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix1}.}
\label{fig:example1}
\end{figure}

\begin{Schunk}
\begin{Sinput}
R> library("dataMaid")
R> data("toyData")
R> toyData
\end{Sinput}
\begin{Soutput}
   var1 var2  var3       var4 var5       var6
1   red    1     a -0.6264538    1 Irrelevant
2   red    1     a  0.1836433    2 Irrelevant
3   red    1     a -0.8356286    3 Irrelevant
4   red    2     a  1.5952808    4 Irrelevant
5   red    2     a  0.3295078    5 Irrelevant
6   red    6     b -0.8204684    6 Irrelevant
7   red    6     b  0.4874291    7 Irrelevant
8   red    6     b  0.7383247    8 Irrelevant
9   red  999     c  0.5757814    9 Irrelevant
10  red   NA     c -0.3053884   10 Irrelevant
11 blue    4     c  1.5117812   11 Irrelevant
12 blue   82     .  0.3898432   12 Irrelevant
13 blue   NA       -0.6212406   13 Irrelevant
14 <NA>  NaN other -2.2146999   14 Irrelevant
15 <NA>    5 OTHER  1.1249309   15 Irrelevant
\end{Soutput}
\end{Schunk}
\begin{Schunk}
\begin{Sinput}
R> clean(toyData)
\end{Sinput}
\end{Schunk}

By default, an \proglang{R} markdown file and a rendered pdf overview
document is produced, saved to the working directory and
opened for immediate inspection. Turning to Figure~\ref{fig:example1},
we see that such a data cleaning output document consists of two
parts. First, an overview of what was done is presented under the
title \textit{Data cleaning summary}. Secondly, each variable in the
dataset is presented in turn using (up to) three tools in the
\textit{Variable list}: A table summarizing key features of the
variable, a figure visualizing its distribution and a
list of flagged issues, if any. For instance, in the \code{numeric}-type variable
\code{var2} from \code{toyData}, \code{clean()} has identified two values that
are suspected to be miscoded missing values (\code{999} and \code{NaN}),
while two values were also flagged as potential outliers that should
be investigated more carefully.
%We can then return to Part
%1, the data cleaning summary, and inspect what sorts of checks were
%performed on this variable. \hl{hm, delete the last sentence? or use
%  different variable? Only these exact two checks were performed}.

Though the \code{clean()} function is very easy to use, it should not be
mistaken to be inflexible. A large number of function arguments allows
for the cleaning overview document to be molded according to the
user's needs. The most commonly used arguments are summarized in
Table~\ref{table.cleanFormals} and they are grouped according to the
part of the cleaning process they influence. In order to understand
this distinction, a glimpse of the inner structure of \code{clean()} is
shown in Figure~\ref{figure:cleanStructure}. Below, we present a few
examples on how to use the arguments from Table \ref{table.cleanFormals}
 to influence the output of a \code{clean()} call.

@@

\smallskip Control input variables and summary\\
\quad \code{useVar} & What variables should be used? & \code{NULL} (corresponding to all variables) \\
\quad \code{ordering} & Ordering of the variables in the data summary (as is or alphabetical) & \code{"asIs"} \\
\quad \code{onlyProblematic} & Should only variables flagged as problematic be included in the \textit{Variable list}? & \code{FALSE} \\
\quad \code{listChecks} & Should an overview of what checks were performed be listed in the \textit{Data cleaning summary}? &  \code{TRUE} \\
\quad \code{preChecks} & What check functions should be called to determine whether a variable is suitable for summarization, visualization and checking? & \code{c("isKey", "isEmpty")} \\

@@

\hline
\end{tabular}
\caption{A selection of commonly used arguments to \code{clean()} separated into the parts they control.}
\label{table.cleanFormals}
\end{table}

% Below, we present a selection of these arguments in
% sections \hl{xx} and \hl{xxx}. We do this under two distinct headlines
% to emphasize that two levels of control are available when using
% \code{clean()}. Either, we can control the \hl{[something] - essentially,
%   this is everything but SVC-functions - } or we can control what
% quality assessments and descriptions each variable are subjected
% to.  \hl{segway to next bit or
%   just something to end the paragraph}

\begin{figure}[tb]
% \begin{description}
% \item[Input] A dataset
% 	\begin{itemize}
% 		\item This should be of type \code{data.frame} or \code{tibble} \hl{or data.table?}
% 	\end{itemize}
% \item[Create contents] For each variable in the dataset, do the following:
% 	\begin{description}
% 		\item[Stage 1:] Pre-checks
% 			\begin{itemize}
% 				\item Is the variable suitable for summarization, visualization and checks?
% 				\begin{description}
% 					\item[Yes:] Go to stage 2.
% 					\item[No:] Move on to the next variable.
% 				\end{description}
% 			\end{itemize}
% 		\item[Stage 2:] SVC-steps
% 			\begin{description}
% 				\item[Summarize] Call \code{summarize()} to produce a summary table describing the variable. What features enter this table depends on the data class of the variable.
% 				\item[Visualize]  Call \code{visualize()} to produce a plot visualizing the distribution of the variable.
% 				\item[Check]  Call \code{check()} to apply quality- and error checks to the variable. What checks are used depends on the data class of the variable.
% 			\end{description}
% 	\end{description}
% \item[Output] Files for a overview document are saved to the disc and possibly also opened.
% 	\begin{itemize}
% 		\item Always a \code{rmarkdown} (.Rmd) file
% 		\item Possible also a html or pdf file
% 	\end{itemize}
% \end{description}

% Define block styles
\tikzstyle{decision} = [diamond, draw, fill=blue!20,
    text width=5.5em, text badly centered, node distance=3cm, inner
    sep=0pt]
\tikzstyle{block} = [rectangle, draw, fill=blue!20,
    text width=6em, text centered, rounded corners, minimum
    height=4em]
\tikzstyle{line} = [draw, -latex']
\tikzstyle{cloud} = [draw, ellipse,fill=red!20, node distance=3cm,
    text width=5em,
    minimum height=2em]

@@

\caption{Schematic illustration of the stages undertaken when running
  \code{clean()}. Each variable is checked for eligibility before
  running \code{summarize()}, \code{visualize()}, and \code{check()}, and the
  resulting \proglang{R} markdown file may be rendered and opened.}
\label{figure:cleanStructure}
\end{figure}

@@

\subsection{Polishing off the arguments}
We begin with an example that is intended as an illustration of how
\code{clean()} might be used in the very first stages of data
cleaning, when the we are uncertain about the complexities of the
errors and how much times should be allocated to data cleaning. At
this stage, what is really needed, is a very rough idea of the
severity of errors in the dataset. In this scenario, we might wish to
obtain a summary document in html format that only contains the
variables with potential problems, and with a limit of, say, maximum
10 printed potential errors for each variable. Also, we can add the
argument \code{replace=TRUE} in order to force \code{clean()} to
overwrite any existing files produced by \code{clean()}.  Using the
\code{toyData} dataset as a guinea pig, we type:

\begin{Schunk}
\begin{Sinput}
R> clean(toyData, output = "html", onlyProblematic = TRUE, maxProbVals = 10,
+    replace = TRUE)
\end{Sinput}
\end{Schunk}

The final rendering of the generated markdown file is controlled by
the \code{render} and \code{openResult} arguments, which both default to
\code{TRUE}. \code{render} determines if the \proglang{R} markdown file
produced should be rendered using the \pkg{rmarkdown} \citep{rmarkdown} package and
\code{openResult} decides whether the output html or pdf file should be
opened. The following command produces an \proglang{R} markdown file
containing the data cleaning summary information, but without
rendering nor opening the markdown file:

\begin{Schunk}
\begin{Sinput}
R> clean(toyData, output="html", render=FALSE, openResult=FALSE)
\end{Sinput}
\end{Schunk}

>>>>

%% almost as usual
\author{Anne Helby Petersen\\Biostatistics\\Department of Public
  Health\\University of Copenhagen \And Claus Thorn Ekstr\o m\\Biostatistics\\Department of Public
  Health\\University of Copenhagen}
\title{\pkg{dataMaid}: your assistant for data documentation in \proglang{R}}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Anne Helby Petersen, Claus Thorn Ekstr\o m} %% comma-separated
\Plaintitle{{dataMaid}: your assistant for data documentation in R} %% without formatting
\Shorttitle{\pkg{dataMaid}: your assistant for data documentation in R} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{Data cleaning and -validation are important steps in any
  data analysis, as the validity of the conclusions from the analysis
  hinges on the quality of the input data. Mistakes in the data can
  arise for any number of reasons, including erroneous codings,
  malfunctioning measurement equipment, and inconsistent data generation
  manuals.  Ideally, a human investigator should go
  through each variable in the dataset and look for potential errors
  --- both in input values and codings --- but that process can be very
  time-consuming, expensive and error-prone in itself.

  We describe an \proglang{R} package, \pkg{dataMaid}, which
  implements an extensive and customizeable suite of quality
  assessment tools that can be applied to a dataset in order to
  identify potential problems in its variables. The results are 
  presented in an auto-generated, non-technical, stand-alone overview
  document intended to be perused by an investigator with an
  understanding of the variables in the data, but not necessarily
  knowledge of \proglang{R}. Thereby, \pkg{dataMaid} aids the dialogue
  between data analysts and field experts, while also providing easy
  documentation of reproducible data cleaning steps and data quality
  control. Moreover, the \pkg{dataMaid} solution changes the data
  cleaning process from the usual ad hoc approach to a systematic,
  well-documented endeavor.  \pkg{dataMaid} also provides a suite of
  more typical \proglang{R} tools for interactive data quality
  assessment and -cleaning, where the data inspections are executed directly 
  in the \proglang{R} console.

  % The \pkg{dataMaid} package is designed to be easily extended with
  % custom user-created checks that are relevant in particular
  % situations. \hl{Already said that above. Either expand on it or
  % delete this.}
}
\Keywords{data cleaning, quality control, \proglang{R}, data documentation}
\Plainkeywords{data cleaning, quality control, R, data documentation} %% without formatting
%% at least one keyword must be supplied

@@

\section{Introduction}
Though data cleaning might be regarded as a somewhat tedious activity,
adequate data cleaning is crucial in any data analysis. With
ever-growing dataset sizes and complexities, statisticians and data
analysts find themselves spending a large portion of their time on
data cleaning and data wrangling. While a computer should generally not
make unsupervised decisions on what should be done to potential
errors in a dataset, it can still be an extremely useful tool in the
data cleaning process. Some errors can be tracked down and flagged by a
computer without further ado, while other types of errors need an analytic
context in order to be identified. Even in this latter case, well-designed
software can aid the process tremendously by providing the human
investigator with the information needed for identifying issues.

But even when tools are available for identifying problems in a dataset, the activity of data cleaning still suffers from a challenge that has recieved increasing attention in the scientitic communities in the later years: Data cleaning is not very straight forward to document and therefore, reproducibility suffers. \pkg{dataMaid}'s most central purpose is to facilitate a data cleaning workflow where documentation is thoroughly integrated rather than a secondary add-on. This is accomplished by centering the data cleaning process around auto-generated data overview reports that summarize the contents of the current version of a dataset and flags potential problems that could be candidates for the next bullet on the cleaning agenda.

A number of \proglang{R} packages made for other pre-analysis tools are
already available, including \pkg{janitor} \citep{janitor}, \pkg{dplyr}
\citep{dplyr}, \pkg{tidyr}\citep{tidyr}, \pkg{data.table} \citep{data.table}, \pkg{DataCombine}
\citep{DataCombine} , \pkg{validate} \citep{validate}, \pkg{assertr} \citep{assertr}, and
\pkg{DataExplorer} \citep{DataExplorer}. These packages focus on
different stages of the pre-analysis work, most of which are not
really data cleaning.  \pkg{janitor} provides tools for data import
with a particular emphasis on the challenges of getting neat data
frames from Microsoft Excel data files. \pkg{dplyr}, \pkg{tidyr}, \pkg{data.table}
and \pkg{DataCombine} go a few steps further by providing a wide array
of extremely powerful tools for data wrangling, including a number of
particularly useful functions for merging and working with very large
datasets. When it comes to actual data cleaning, however, the options
are fewer.  \pkg{validate} (and the similar packages \pkg{editrules}
\citep{editrules} and \pkg{deducorrect} \citep{deducorrect} from the
same authors) does offer a few tools for identifying and
auto-correcting errors in a dataset, but, as the name suggests, the
focus is on internal validity rather than general data cleaning. In
practice, this means that quite elegant tools for, e.g., linear
restraints among the variables in a dataset can be applied, but
looking for potentially miscoded missing values is not really
feasible. The main difference between these two challenges is the
direction in which the data is inspected: While linear constraints
work observation-wise with no ambiguity, determining whether or not something
is a miscoded missing value often requires knowledge about the full
variable (e.g. range or data type), and thus it should be performed
variable-wise. \pkg{validate} does not allow for user-defined extensions
of the latter type, thereby limiting its data cleaning potential.
Automatic data correction functions are also provided by
\pkg{validate} which we consider to be quite a dangerous cocktail: all
power is given to the the computer with no human supervision, and
investigators are less likely to make an active, case-specific choice
regarding the handling of the potential errors.  Finally, no tools
have been made to easily document exactly which checks and preliminary
results were used in the data cleaning process. \hl{Something about \pkg{assertr} here.}

@@

The \pkg{dataMaid} approach to data cleaning, -quality assessment and -documentation is
governed by two fundamental paradigms. First of all, there is no need
for data cleaning to be an ad hoc procedure. Often, we have a very
clear idea of what flags are raisable in a given dataset before we
look at it, as we were the ones to produce it in the first place. This
means that data cleaning can easily be a well-documented,
well-specified procedure. In order to aid this paradigm, \pkg{dataMaid}
provides easy-to-use, automated tools for data quality assessment in
\proglang{R} \citep{R} on which data cleaning decisions can be made. This quality
assessment is presented in an auto-generated overview document,
readable by data analysts and field experts alike, thereby also
contributing to an inter-field dialogue about the data at
hand. Oftentimes, e.g., distinguishing between faulty codings of a
numeric value and unusual, but correct, values requires
problem-specific expertise that might not be held by the data
analyst. Hopefully, having easy access to data descriptions through
\pkg{dataMaid} will help this necessary knowledge sharing.

@@

This manuscript is structured as follows: First, in Section
\ref{sec:usingdataMaid}, we introduce the main representative of the first
paradigm, namely the \code{clean()} function, which generates data
overview documents. In the \pkg{dataMaid} package, we have
provided a number of default checks that cover the data
cleaning challenges, we find to be most common. Next, in Section
\ref{sec:interactiveCleanR}, we present the interactive mode of \pkg{dataMaid}, as motivated
by the second paradigm above. But, as any data analyst knows,
every dataset is different, and some datasets might include problems
that cannot be detected by our data checking functions. In Section
\ref{sec:extending}, we turn to the question of how \pkg{dataMaid} extensions
can be made, such that they are integrable with the \code{clean()}
function and with the other tools available in \pkg{dataMaid}.  At last,
in Section \ref{sec:specificExamples}, we discuss a number of examples of
specific data cleaning challenges and how \pkg{dataMaid} can be used to
solve them.

@@

\section{Creating a data overview report}
\label{sec:usingdataMaid}

The \code{makeDataReport()} function is the primary workhorse of \pkg{dataMaid} and
this is the only function needed if a standard battery
of tests are used to generate data reports. The data report itself is an overview document, intended for reading
by humans, in either pdf or html format. Appendix \ref{sec:appendix1}
provides an example of a data report, produced by
calling \code{makeDataReport()} on the dataset \code{toyData} available in
\pkg{dataMaid}. The first two pages of this data report are
shown in Figure~\ref{fig:example1}. \code{toyData} is a very
small ($15$ observations of $6$ variables), artificial dataset which was created to
illustrate the main capabilities of \pkg{dataMaid}. The following
commands load the dataset and produce the report:

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=2]{dataMaid_toyData.pdf}}
\frame{\includegraphics[width=7.5cm,page=3]{dataMaid_toyData.pdf}}
%\includepdf[pages={2}, pagecommand={}]{dataMaid_testData.pdf}
\end{center}
\caption{Example output from running \code{makeDataReport()} on the \code{toyData}
  dataset. First a summary of the full dataset is given and then
  type-dependent information on each variable is given in a table and
  a graph. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix1}.}
\label{fig:example1}
\end{figure}

\begin{Schunk}
\begin{Sinput}
R> library("dataMaid")
R> data("toyData")
R> toyData
\end{Sinput}
\begin{Soutput}
   var1 var2  var3       var4 var5       var6
1   red    1     a -0.6264538    1 Irrelevant
2   red    1     a  0.1836433    2 Irrelevant
3   red    1     a -0.8356286    3 Irrelevant
4   red    2     a  1.5952808    4 Irrelevant
5   red    2     a  0.3295078    5 Irrelevant
6   red    6     b -0.8204684    6 Irrelevant
7   red    6     b  0.4874291    7 Irrelevant
8   red    6     b  0.7383247    8 Irrelevant
9   red  999     c  0.5757814    9 Irrelevant
10  red   NA     c -0.3053884   10 Irrelevant
11 blue    4     c  1.5117812   11 Irrelevant
12 blue   82     .  0.3898432   12 Irrelevant
13 blue   NA       -0.6212406   13 Irrelevant
14 <NA>  NaN other -2.2146999   14 Irrelevant
15 <NA>    5 OTHER  1.1249309   15 Irrelevant
\end{Soutput}
\end{Schunk}
\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData)
\end{Sinput}
\end{Schunk}

By default, an \proglang{R} markdown file and a rendered pdf overview
document is produced, saved to the working directory and
opened for immediate inspection. Turning to Figure~\ref{fig:example1},
we see that such a data report consists of three
parts. First, an overview of what was done is presented under the
title \textit{Data report overview}. Secondly, an index listing each variable along with an indication of whether it was found to be problematic or not is provided. Thirdly, each variable in the
dataset is presented in turn using (up to) three tools in the
\textit{Variable list}: A table summarizing key features of the
variable, a figure visualizing its distribution and a
list of flagged issues, if any. For instance, in the \code{numeric}-type variable
\code{var2} from \code{toyData}, \code{makeDataReport()} has identified two values that
are suspected to be miscoded missing values (\code{999} and \code{NaN}),
while two values were also flagged as potential outliers that should
be investigated more carefully.

Though the \code{makeDataReport()} function is very easy to use, it should not be
mistaken to be inflexible: By using the optional arguments of the function, 
both the contents and the look of the data report can be molded 
according to the user's needs. The most commonly used arguments are summarized in
Table~\ref{table.cleanFormals} and they are grouped according to the
part of the data assesment and report generation they influence. In order to understand
this distinction, a glimpse of the inner structure of \code{makeDataReport()} is
shown in Figure~\ref{figure:cleanStructure}. Below, we present a few
examples on how to use the arguments from Table \ref{table.cleanFormals}
 to influence the output of a \code{makeDataReport()} call.

@@

\smallskip Control input variables and summary\\
\quad \code{useVar} & What variables should be used? & \code{NULL} (corresponding to all variables) \\
\quad \code{ordering} & Ordering of the variables in the data summary (as is or alphabetical) & \code{"asIs"} \\
\quad \code{onlyProblematic} & Should only variables flagged as problematic be included in the \textit{Variable list}? & \code{FALSE} \\
\quad \code{listChecks} & Should an overview of what checks were performed be listed in the \textit{Data report overview}? &  \code{TRUE} \\
\quad \code{preChecks} & What check functions should be called to determine whether a variable is suitable for summarization, visualization and checking? & \code{c("isKey", "isEmpty")} \\

@@

\hline
\end{tabular}
\caption{A selection of commonly used arguments to \code{makeDataReport()} separated into the parts they control.}
\label{table.cleanFormals}
\end{table}

\begin{figure}[tb]
% Define block styles
\tikzstyle{decision} = [diamond, draw, fill=blue!20,
    text width=5.5em, text badly centered, node distance=3cm, inner
    sep=0pt]
\tikzstyle{block} = [rectangle, draw, fill=blue!20,
    text width=6em, text centered, rounded corners, minimum
    height=4em]
\tikzstyle{line} = [draw, -latex']
\tikzstyle{cloud} = [draw, ellipse,fill=red!20, node distance=3cm,
    text width=5em,
    minimum height=2em]

@@

\caption{Schematic illustration of the stages undertaken when running
  \code{makeDataReport()}. Each variable is checked for eligibility before
  running \code{summarize()}, \code{visualize()}, and \code{check()}, and the
  resulting \proglang{R} markdown file may be rendered and opened.}
\label{figure:cleanStructure}
\end{figure}

@@

\subsection{Polishing off the arguments}
We begin with an example that is intended as an illustration of how
\code{makeDataReport()} might be used in the very first stages of data
cleaning, when the we are uncertain about the complexities of the
errors and how much time should be allocated to data cleaning. At
this stage, what is really needed, is a very rough idea of the
severity of errors in the dataset. In this scenario, we might wish to
obtain a summary document in html format that only contains the
variables with potential problems, and with a limit of, say, maximum
10 printed potential errors for each variable. Also, we can add the
argument \code{replace=TRUE} in order to force \code{makeDataReport()} to
overwrite any existing files produced by \code{makeDataReport()}.  Using the
\code{toyData} dataset as a guinea pig, we type:

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, output = "html", onlyProblematic = TRUE, maxProbVals = 10,
+    replace = TRUE)
\end{Sinput}
\end{Schunk}

The final rendering of the generated markdown file is controlled by
the \code{render} and \code{openResult} arguments, which both default to
\code{TRUE}. \code{render} determines if the \proglang{R} markdown file
produced should be rendered using the \pkg{rmarkdown} \citep{rmarkdown} package and
\code{openResult} decides whether the output html or pdf file should be
opened. The following command produces an \proglang{R} markdown file
containing the information needed for generating a data report, but without
rendering nor opening the markdown file:

\begin{Schunk}
\begin{Sinput}
R> makeDataReport(toyData, output="html", render=FALSE, openResult=FALSE)
\end{Sinput}
\end{Schunk}

\begin{table}
\centering
\begin{tabular}{p{0.35\linewidth} p{0.3\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth}
 p{0.01\linewidth} p{0.01\linewidth}}
  \hline
& Description &  \multicolumn{7}{c}{Variable classes} \\ \smallskip
 & &  C & F & I & L & B & N & D\\
  \hline \smallskip
  \textbf{\code{summaryFunction}s}  \smallskip \\
  \quad \code{centralValue} & Compute median or mode &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \quad \code{countMissing} & Compute proportion of missing observations &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \quad \code{minMax} & Find minimum and maximum values &   &  & $\times$ & &  & $\times$ & $\times$  \\
  \quad \code{quartiles} & Compute 1st and 3rd quartiles &    &  & $\times$ & &  & $\times$ & $\times$ \\
  \quad \code{uniqueValue} & Count number of unique values &   $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \quad \code{variableType} & Data class of variable & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \smallskip \\
 \textbf{\code{visualFunction}s} \smallskip \\
  \quad \code{basicVisual} & Histograms and barplots using base \proglang{R} graphics &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \quad \code{standardVisual} & Histograms and barplots using \pkg{ggplot2} &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \smallskip \\
 \textbf{\code{checkFunction}s} \smallskip \\
 \quad \code{identifyCaseIssues} & Identify case issues &  $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyLoners} & Identify levels with $<$ 6 obs. & $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyMissing} & Identify miscoded missing values &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &  \\
 \quad \code{identifyNums} & Identify misclassified numeric or integer variables & $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyOutliers} & Identify outliers &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \code{identifyOutliersTBStyle} & Identify outliers (Turkish Boxplot style) &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \code{identifyWhitespace} & Identify prefixed and suffixed whitespace &  $\times$ & $\times$ & & $\times$ & & &  \\
 \quad \code{isCPR} & Identify Danish CPR numbers & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &$\times$   \\
 \quad \code{isEmpty} & Check if the variable contains only a single value & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
 \quad \code{isKey} & Check if the variable is a key & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$\smallskip   \\
 \hline
\end{tabular}
\caption{List of all summary functions (used in the summary table for
  each variable in the output), visual functions (used for  visualization of each variable), and
  check functions (used for data checks for each variable) currently implemented in \pkg{dataMaid}. The variable
  classes C, F, I, L, B, N, and D, refer to \code{character}, \code{factor},
  \code{integer}, \code{labelled}, \code{logical} (boolean), \code{numeric}, and \code{Date}, respectively.}
\label{table.SVCfunctions}
\end{table}

We will now move on to discuss how not only the structure of the
data assesment steps is manipulated, but also its very contents. This is
done through the summarize/visualize/check (SVC) steps, as illustrated
in Figure \ref{figure:cleanStructure}.  \pkg{dataMaid} uses three
different types of functions for performing these steps, namely
\code{summaryFunction}s, \code{visualFunction}s and
\code{checkFunction}s.  By default, \code{clean()} runs selected summary,
visualization and check functions on each variable in the dataset, and
the exact choice of these functions depends on the classes of the
variables. For instance, though detection of outlier values might be
interesting for numerical variables, it holds little meaning for
factor variables, and therefore, numerical and factor variables need
different checks. Table~\ref{table.SVCfunctions} lists all available
summarize/visualize/check functions, but we can also use the
\code{allSummaryFunctions()}, \code{allVisualFunctions()}, and
\code{allCheckFunctions()} functions in \pkg{dataMaid} to print overview
lists in \proglang{R}. For example, the implemented
\code{summaryFunction}s are:
